---
title: "Chapter 8: Your Own Functions!"
author: "Tyson S. Barrett"
date: "Summer 2017"
institute: "Utah State University"
fontsize: 11pt
output:
  beamer_presentation:
    theme: "metropolis"
    toc: true
    fig_width: 4.5
    fig_height: 3.5
    fig_caption: true
    df_print: default
    keep_tex: true
    incremental: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

## Introduction

**User-defined functions** --- functions you create yourself --- are an important tool.

We will show how you can make your own functions.

## What is a Function Anyway?

It is a special `R` object that performs something on another object or the environment.

For example, 

- `ggplot()` takes a `data.frame` object and variable names and produces a plot.
- `lm()` takes a formula, a `data.frame` and performs a statistical model.
- `gather()` takes a `data.frame` and produces another `data.frame` in another form


# Creating a Function
## Creating a Function

It all starts with the function `function()`.
```{r, eval=FALSE}
myfunction <- function(arguments){
  stuff = that(you)
  want = it + to + do + 4
  you
}
```
After `function()` we use `{` and `}`. Everything in between is what you want the function to do.

All you need to do is run the function and you can use it in the `R` session.

# Named Functions
## Named Functions

These are functions that you assign a name to.
```{r}
mean2 <- function(x){
  n <- length(x)
  m <- (1/n) * sum(x)
  return(m)
}
```
Now, `mean2()` is a function you can use. 

*What does this function do?*

## Named Functions

```{r}
v1 <- c(1,3,2,4,2,1,2,1,1,1)   ## vector to try
```

Let's give it a try using the vector `v1`

```{r}
mean2(v1)                      ## our function
mean(v1)                       ## the base R function
```

## Named Functions

For practice, we will write one more function. Let's make a function that takes a vector and gives us the N, the mean, and the standard deviation.
```{r}
important_statistics <- function(x, na.rm=FALSE){
  N  <- length(x)
  M  <- mean(x, na.rm=na.rm)
  SD <- sd(x, na.rm=na.rm)
  
  final <- c(N, M, SD)
  return(final)
}
```

## Named functions

One of the first things you should note is that we included a second argument in the function seen as `na.rm=FALSE` (you can have as many arguments as you want within reason). 

- This argument has a default that we provide as `FALSE` 
- We take what is provided in the `na.rm` and give that to both the `mean()` and `sd()` functions. 
- Finally, you should notice that we took several pieces of information and combined them into the `final` object and returned that.

## Named Functions

Let's try it out with the vector we created earlier.
```{r}
important_statistics(v1)
```

## Named Functions

Looks good but we may want to change a few aesthetics. In the following code, we adjust it so we have each one labeled.
```{r}
important_statistics2 <- function(x, na.rm=FALSE){
  N  <- length(x)
  M  <- mean(x, na.rm=na.rm)
  SD <- sd(x, na.rm=na.rm)
  
  final <- data.frame(N, "Mean"=M, "SD"=SD)
  return(final)
}
important_statistics2(v1)
```
We will come back to this function and use it in some loops and see what else we can do with it.



# Anonymous Functions
## Anonymous Functions

Sometimes it is not worth saving a function but want to use it, generally within loops.

```{r, eval=FALSE}
function(x) thing(that, you, want, it, to, do, with, x)
```

We will show a few of these in the looping section (although they are identical in nature to named functions, they just aren't named)


# Why Write Your Own?
## Why Write Your Own?

Several reasons exist.

- Looping
- Adjusting output
- Performing a special function
- Other customization

We are going to talk about looping in depth.

## Looping

Writing your own functions for looping is very common and practical.

Loops are things that are repeated.

For example:

- We may want a certain statistic (like a mean) for every continuous variable in the data set.
- We may want to remove `999` from every variable in a data set.
- We may want to change variable types of certain variables across the whole data set.


## Looping

Examples in `R` include:

- `for` loops
- the `apply` family of functions

## For Loops

```{r, eval=FALSE}
for (i in 1:10){
  mean(data[, i])
}
```


## For Loops

Another example:
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
data = read.csv("~/Dropbox/Teaching/R for Social Sciences/Data/WideFormat_TheOffice.csv") %>%
  select(Prod1, MentalApt, PhysApt, Income, Children, SubsUse, Ment1, Ment2)
thing = list()
for (i in 1:8){
  thing[[i]] = cbind(mean(data[, i]), sd(data[, i]))
}
```

## For Loops

```{r}
thing
```

## For loops

I like for loops. They are easy to understand and fiddle with, after some practice.

However, they used to be slow in `R` and so they have a bad reputation.

## The apply family

There are several `apply` functions in `R` that do loops for you.

- `apply()`
- `sapply()`
- `lapply()`
- `tapply()`

## sapply

Produces a vector based on the function that it is repeating.

Both do the same thing here.

```{r, eval=FALSE}
for (i in 1:10){
  mean(data[, i])
}
sapply(1:10, function(i) mean(data[, i]))
```

## sapply

Can also just provide the `data.frame` and it assumes you want the function (in this case `mean()`) applied to each variable.
```{r}
sapply(data, mean)
```

## lapply

Produces a list. Just like `sapply()` it takes a `data.frame` and a function and applies it across the variables.

```{r}
lapply(data, mean)
```


## tapply

Is a bit different than the rest. It doesn't do much in terms of looping necessarily (although you can have it do that). Instead, it applies a function based on a grouping variable. With the `tidyverse`, however, this is not often used any more.

```{r}
tapply(data$Income, data$Children, mean)
```

*How could you do this in the `tidyverse` framework?*


# Loops with User-Defined Functions
## Loops with User-Defined Functions

Going back to our `important_statistics2()` function:

```{r}
important_statistics2 <- function(x, na.rm=FALSE){
  N  <- length(x)
  M  <- mean(x, na.rm=na.rm)
  SD <- sd(x, na.rm=na.rm)
  
  final <- data.frame(N, "Mean"=M, "SD"=SD)
  return(final)
}
```
Let's put it in a loop.

## Loops with User-Defined Functions

```{r}
lapply(data, important_statistics2)
```

## Loops with User-Defined Functions

```{r}
sapply(data, important_statistics2)
```

## Loops with User-Defined Functions

It applied our function across the variables in the `data.frame`. So we can easily get information we want. 

This was a very simplified version of how I created the `table1()` function in `furniture`.

# Conclusions
## Conclusions

Writing your own functions takes time and practice but it can be a worthwhile tool in using `R`.

I recommend you start simple and start soon.

Ultimately, you can make your own group of functions you use often and create a package for it so others can use them too :)



