---
title: "Miscellaneous Topics"
author: "Tyson S. Barrett"
date: "Summer 2017"
institute: "Utah State University"
fontsize: 10pt
output:
  beamer_presentation:
    theme: "metropolis"
    toc: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
    df_print: default
    keep_tex: true
    incremental: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
## Good Quote

> "All the cool kids use R."
>
> --- Anonymous

## Introduction
We need to cover a few additional topics, as well as review some topics:

1. Loops (More Concrete Examples)
2. RMarkdown
3. Finding Relevant Packages
4. Modeling Summary
5. Final Quiz

# Loops
## Loops

This is a very important tool to use in data cleaning.

\vspace{25pt}

##### Example: You have `999` as a placeholder throughout your data. How can you quickly change those to `NA`?

## Loops
Here's one way:
```{r, eval=FALSE}
library(furniture)
df[] <- lapply(df, washer, 999)
```

And another:
```{r, eval=FALSE}
df[] <- lapply(df, function(x) ifelse(x == 999, NA, x))
```

\vspace{20pt}

Note: the `[]` after the `df` forces `lapply()` to keep it in a data.frame :)

## Loops
##### Another example: You need to recode the 5th, 10th, and 20th through 30th variables. How can you do that with a loop?

## Loops
Here's one way:
```{r, eval=FALSE}
library(forcats)
df[, c(5,10,20:30)] <- lapply(df[, c(5,10,20:30)], fct_recode, 1 = "5", 2 = "4", 4 = "2", 5 = "1")
```

And another:
```{r, eval=FALSE}
library(forcats)
for (i in c(5,10,20:30)){
  df[, i] = fct_recode(df[, i], 1 = "5", 2 = "4", 4 = "2", 5 = "1")
}
```

# RMarkdown
## RMarkdown

So much to do! But focus on the basics :)

1. The key to the whole thing is that you can use regular text and `R` code together.
2. The `R chunks` are very flexible so that you usually can get the results to look the way you want.
3. Markdown is a markup language but is not a WYSIWYG.[^1] But, you can always knit to see how it is looking.
4. Practice with it. Try things out.

[^1]: What You See Is What You Get

## RMarkdown
The text can be any text you would normally put in a document.

Code chunks can be any functioning `R` code.

Check out the following links for more information: 

1. [RStudio and RMarkdown](http://rmarkdown.rstudio.com/lesson-1.html)
2. [Nice Youtube Tutorial](https://www.youtube.com/watch?v=DNS7i2m4sB0)
3. [Other Nice, More Indepth Tutorial](https://www.youtube.com/watch?v=cWJzjHh_3kk)

# Finding Relevant Packages
## Finding Relevant Packages

Much of this results from two things:

1. Experience: you find packages that you like and trust over time
2. Google: The top results in Google are often good ones (at least well used).

## Finding Cronbach's Alpha

Example of finding a package for cronbach's alpha.

# Modeling Summary
## Modeling Summary

I wanted to provide you with a quick summary of packages and functions needed for different model types. 

```{r, echo=FALSE}
models = cbind("Model" = c("T-Test", "ANOVA", "Linear", "Logistic", "Poisson", "GEE", "Mixed Effects (MLM)", "SEM"),
               "Package" = c("", "", "", "", "", "gee or geepack", "lme4", "lavaan"),
               "Function" = c("t.test()", "aov()", "lm()", "glm() with 'binomial'", "glm() with 'poisson'", "gee() or geeglm()", "lmer()", "sem() or cfa()"))
knitr::kable(models)
```



# Final Quiz
## Final Quiz

Let's see if you are comfortable with the following scenarios/lines of code.

## 1. Data Manipulation

You want to do three things:

1. Create a binary depression variable based on the level of the continuous depression variable (above a value of 10),
2. Filter out those that have no depression (level 0 of the new depression variable),
3. Find the mean, sd, and range of the productivity variable by sex.

## 2. Reshape That

Your data is in wide format. It is a longitudinal data set, with two observations per individual. What steps do you take to reshape it into long format?

## 3. Visualize It

Your data is again in wide format. It is longitudinal, with two observations per individual.

You want to create:

1. A scatterplot comparing mental aptitude and productivity at time one,
2. A line graph showing the means of productivity by sex at each time point,
3. Someone sends you the following code and you want to know what it does.

```{r, eval=FALSE}
ggplot(df, aes(x = Sex, y = Productivity, color = Sex)) +
  geom_boxplot() +
  scale_color_manual(values = c("red", "not_red")) +
  theme(legend.position = "bottom")
```


## 4. Strut Yo' Stuff (i.e. Modeling)

If you have repeated measures data, what format does you data need to be in to analyze it with either RM-ANOVA or Mixed Effects?

## 5. RMarkdown

How do you create a new `rmarkdown` file?

In `rmarkdown` how can you insert inline r syntax?


# Closing Remarks
## Thanks!

Thanks for participating. I hope this was a great start for you in using `R` throughout your careers. 

Let me know if I can help in the future with `R` or stat related things.


